Types for R, Empirically Designed and Evaluated

### 2020-05-07 paper/main.tex
Improve usability discussion

@@

\subsection{contractr}
\label{sec:contractr}
\emph{contractr} is an R package that implements our type system for the R
package ecosystem. It adds contracts to package functions for type-checking
arguments and results on function calls. On contract failure, it reports a type
mismatch warning with the package name, function name, argument name, argument
position, expected type signature, actual type signature of the value observed,
and a stack trace for the called function. \emph{contractr} works by modifying
function definition to insert a call to the type-checking function. When the
modified function is called, the type-checking function is invoked on the
modified function's arguments and return value.

\emph{contractr} is implemented in 434 R LOC and 2358 C++ LOC. The core
implementation is in C++ to keep the runtime overhead low. It been designed and
tested with GNU R-3.5.0 but it also works with recent versions of R. It has been
hardened with a battery of 400 test cases. We have used it extensively during
the course of this work; firstly, for sanity checking of XXX type signatures
generated for the XXX packages during the development phase, and secondly, for
assessing the quality of these type signatures on XXX packages during the
evaluation phase.

%
%
\subsubsection{Type Declarations}
Type declarations can be made available to \emph{contractr} in three ways: as
part of its internal database of types (which already has type signatures for
our corpus of 500 packages), as part of a designated \emph{TYPEDECLARATION} file
supplied by the package author and installed along with the package, or through
a user-level API function insert\_contract that lets the user insert contract to
a custom function with the type declaration provided as a string argument.

%
%
\subsubsection{Challenges}
Retrofitting a robust and efficient type-system as an external package in R has
been a significant undertaking. We had to contend with the complex design
choices of R to make it work without surprises for the end-users. We describe
two key challenges that we faced in the development of \emph{contractr}.
Firstly, we faced the problem of type-checking arguments in a non-strict
language while retaining the non-strict semantics. When a function is called in
R, parameters are bound to unevaluated code thunks called
promises~\cite{oopsla19}, instead of values which can be immediately
type-checked. Naively forcing promises at function call to obtain a value for
type-checking would violate the language semantics, leading to incorrect
results. To preserve the original semantics, \emph{contractr} modifies the
promise by wrapping the unevaluated promise expression in a call to its
type-checker along with the context necessary to associate the promise to the
function and its parameter (for failure messages). The contract checking happens
when the promise is evaluated, either inside the function or many calls deep.
This works, but with one wrinkle. GNU R has a bytecode compiler that can
sometimes optimize away promises; in those cases, \emph{contractr} receives
values that can be immediately type-checked. Non-strict semantics dictate that
any computation related to the argument should happen on the first use of the
argument. So type-checking at this point and issuing an error message would
violate the non-strict semantics; if this argument is never used, it would be
incorrect to type-check it at the beginning of call. To make this work,
\emph{contractr} mutates the function parameter binding to a promise that wraps
the argument value in a call to the type-checker, as in the previous case.
Secondly, we faced the problem of type-checking function return value. In R, the
result of a function call is the result of evaluating the last expression in the
function body; so an explicit return call is absent from most function
definitions. Thus, there could be many potential sub-expressions in the function
body which need to be type-checked aginst the return type. To address this,
\emph{contractr} registers the type-checker to be called on the return value
through a function exit hook. This hook is executed in the function call's
environment after it has executed. While this works, we have to contend with yet
another wrinkle, R interpreter can perform longjumps in its C implementation,
which causes active R function calls on the stack to be discarded. When they are
discarded, their exit hooks are called, which in this case calls the registered
type-checker. But, these functions are in the middle of an active computation,
so they don't have a return value to type-check. \emph{contractr} deals with
this problem by allocating a unique sentinel object which serves as the return
value for calls that are discarded. The function call exit hook does not call
the type-checker if this unique value happens to be the return value of the
call.

%
%
\subsubsection{Usability}
Through this package, we intend to provide a clear path for the R developers to
add types to untyped package code. Hence, we have paid a lot of attention to
usability in its design. Firstly, the package enables type-checking for all
packages with type signatures, automatically, on importing, offering a seamless
and hassle-free experience for both package developers and end-users. A mere
library(\emph{contractr}) is enough to insert contracts, both, in packages
pre-loaded in the user's workspace and in packages that will be loaded
eventually. For this, \emph{contractr} relies heavily on R's reflective and
dynamic capabilities. Upon loading \emph{contractr} scans all the pre-loaded
packages in the user's workspace and inserts contracts in functions for which
type signatures are available. For all other packages installed on user's
machine but not yet loaded in the workspace, contractr sets up package load
hooks. The hooks get executed when these packages get loaded and insert
contracts to the package functions. The ensures that users do not fiddle
manually with the \emph{contractr} API to enable type-checking for packages
loaded eventually. Furthermore, \emph{contractr} automatically removes contracts
from all the package functions and restores them to their original state when it
is unloaded, an operation that is rarely performed but is possible, nonetheless.
Secondly, we enable package authors to supply type declarations to
\emph{contractr} for their package functions, without requiring any modification
to package code. This is accomplished by providing type declarations inside
roxygen2 function comment blocks. roxygen2 is an R package that enables authors
to add documentation to R functions in the form of organized plaintext sections
which are automatically exported to R's latex style custom documentation format.
roxygen2 is used by … R packages in CRAN. It also provides an extension API for
processing metadata in custom documentation tag sections. A package can register
a method to be called when this tag is encountered. The metadata for that tag
section is made available to the registrant before package build and
installation steps. \emph{contractr} uses this API to implement a @type tag. Our
roxygen2 hook extracts type declarations from these sections from all the
documented functions of the package and stores then in a TYPEDECLARATION file
inside the package folder. This files is copied as is on package installation
and automatically picked up \emph{contractr} when the two packges are loaded.
Here is an example of how a roxygen code block looks like with our type
declaration:

This technique enables the function and its type signature to coexist next to
each other, where they are more likely to remain synchronized. It also enables
package authors to use our type system without modifying their code. Thirdly, we
provide a very expressive API and package switches to the users, covering a
variety of uses cases. During development, developers can explicitly insert
contracts by supplying the type declaration as a string argument to the
insert\_contract API function. Contracts can be removed by calling
remove\_contract on a function. Contracts can be selectively enabled or disabled
for code blocks by wrapping them in calls to ignore\_contracts and
capture\_contracts functions. This enables developers and users to selectively
type check sections of code. Furthermore, these functions return a data frame
that contains all the information about failed and successful contract
assertions in the wrapped code blocks. Multiple type-checking failures can
introduce a lot of noise in the program output. To alleviate this problem, users
can set a flag, \emph{contractr}.severity, to “silence” which suppresses the
failure messages. The contract assertions can then be obtained as a data frame
through a call to get\_contracts at any program point. Setting this flag to
“error” will turn the type-checking failure warnings to errors.

>>>>

\newcommand{\contractr}{\emph{contractr}\xspace} % contractr
\newcommand{\roxygen}{\emph{roxygen2}\xspace} % roxygen

@@

\subsection{contractr}
\label{sec:contractr}
\contractr is an R package that implements our type system for the R
package ecosystem. It adds contracts to package functions for type-checking
arguments and results on function calls. On contract failure, it reports a type
mismatch warning with the package name, function name, argument name, argument
position, expected type signature, actual type signature of the value observed,
and a stack trace for the called function. \contractr works by modifying
function definition to insert a call to the type-checking function. When the
modified function is called, the type-checking function is invoked on the
modified function's arguments and return value.

\contractr is implemented in 434 R LOC and 2358 C++ LOC. The core
implementation is in C++ to keep the runtime overhead low. It been designed and
tested with GNU R-3.5.0 but it also works with recent versions of R. It has been
hardened with a battery of 400 test cases. We have used it extensively during
the course of this work; firstly, for sanity checking of XXX type signatures
generated for the XXX packages during the development phase, and secondly, for
assessing the quality of these type signatures on XXX packages during the
evaluation phase.

%
%
\subsubsection{Type Declarations}
Type declarations can be made available to \contractr in three ways: as
part of its internal database of types (which already has type signatures for
our corpus of 500 packages), as part of a designated \emph{TYPEDECLARATION} file
supplied by the package author and installed along with the package, or through
a user-level API function insert\_contract that lets the user insert contract to
a custom function with the type declaration provided as a string argument.

%
%
\subsubsection{Challenges}
Retrofitting a robust and efficient type-system as an external package in R has
been a significant undertaking. We had to contend with the complex design
choices of R to make it work without surprises for the end-users. We describe
two key challenges that we faced in the development of \contractr.
Firstly, we faced the problem of type-checking arguments in a non-strict
language while retaining the non-strict semantics. When a function is called in
R, parameters are bound to unevaluated code thunks called
promises~\cite{oopsla19}, instead of values which can be immediately
type-checked. Naively forcing promises at function call to obtain a value for
type-checking would violate the language semantics, leading to incorrect
results. To preserve the original semantics, \contractr modifies the
promise by wrapping the unevaluated promise expression in a call to its
type-checker along with the context necessary to associate the promise to the
function and its parameter (for failure messages). The contract checking happens
when the promise is evaluated, either inside the function or many calls deep.
This works, but with one wrinkle. GNU R has a bytecode compiler that can
sometimes optimize away promises; in those cases, \contractr receives
values that can be immediately type-checked. Non-strict semantics dictate that
any computation related to the argument should happen on the first use of the
argument. So type-checking at this point and issuing an error message would
violate the non-strict semantics; if this argument is never used, it would be
incorrect to type-check it at the beginning of call. To make this work,
\contractr mutates the function parameter binding to a promise that wraps
the argument value in a call to the type-checker, as in the previous case.
Secondly, we faced the problem of type-checking function return value. In R, the
result of a function call is the result of evaluating the last expression in the
function body; so an explicit return call is absent from most function
definitions. Thus, there could be many potential sub-expressions in the function
body which need to be type-checked aginst the return type. To address this,
\contractr registers the type-checker to be called on the return value
through a function exit hook. This hook is executed in the function call's
environment after it has executed. While this works, we have to contend with yet
another wrinkle, R interpreter can perform longjumps in its C implementation,
which causes active R function calls on the stack to be discarded. When they are
discarded, their exit hooks are called, which in this case calls the registered
type-checker. But, these functions are in the middle of an active computation,
so they don't have a return value to type-check. \contractr deals with
this problem by allocating a unique sentinel object which serves as the return
value for calls that are discarded. The function call exit hook does not call
the type-checker if this unique value happens to be the return value of the
call.

%
%
\subsubsection{Usability}
With \contractr, we aim to provide a hassle-free path to the R developers for
adding types to untyped package code. Hence, we have paid a lot of attention
towards usability in its design. We discuss three such usability features.
\begin{itemize}

\item \contractr enables automatic type-checking on importing without
  any user intervention. Just running \code{library(contractr)} R code is enough
  to insert contracts, both, in packages pre-loaded in the user's workspace, and,
  in packages that will be loaded eventually. To achieve this automation,
  \contractr relies heavily on R's reflective and dynamic capabilities. On
  loading, \contractr scans all the pre-loaded packages in the user's workspace
  and inserts contracts in functions for which type signatures are available. For
  all other packages installed on user's machine but not yet loaded in the
  workspace, \contractr sets up package load hooks which are executed by R when
  those packages are loaded. The hook registered by \contractr insert contracts to
  those package's functions. Thus, developers and users do not have to call the
  \contractr API to enable type-checking for packages at any point of their
  programming workflow. Furthermore, \contractr automatically removes contracts
  from all the package functions and restores them to their original state when it
  is unloaded, an operation that is rarely performed, but useful in interactive
  settings.

\item \contractr enables package authors to supply type declarations for their
  package without requiring package code modifications. Type declarations can be
  written alongside a \code{@type} section inside \roxygen function comment
  blocks. \roxygen is an R package that enables authors to add documentation to
  R functions in the form of organized plaintext sections which are
  automatically exported to R's latex style custom documentation format.
  \roxygen is a widely adopted package, used by over XXX R packages in CRAN. It
  provides an extension API for adding custom documentation tag sections in
  comment blocks. The sections are processed by \roxygen before package build
  step and a registered method for each custom tag section is invoked with the
  section data and documented object. \contractr uses this mechanism to register
  a hook for a custom \code{@type} tag. This hook parses the type declarations
  from the \code{@type} sections, extracts the function name from the documented
  function definition, and stores them in a TYPEDECLARATION file inside the
  package folder. This files is copied verbatim on package installation and
  picked up \contractr when the package is loaded. Here is an example of how a
  roxygen code block looks like with our type declaration:

  This feature enables a function and its type signature to coexist next to each
  other, where they are more likely to remain synchronized. Furthermore, the
  \contractr hook can also add the function's type declaration to its
  documentation, which seamlessly integrate our type system with the existing R
  tooling.

\item \contractr provides a very expressive API to the users, covering a variety
  of uses cases. During interactive development, developers can explicitly
  insert contracts by supplying the type declaration as a string argument to the
  \code{insert\_contract} API function. Conversely, contracts can be removed by
  calling the \code{remove\_contract} API on functions. Contracts can be
  selectively enabled or disabled for code blocks by wrapping them in calls to
  \code{ignore\_contracts} and \code{capture\_contracts} functions. This enables
  selective type-checking of code section during the development phase.
  Furthermore, these functions return a data frame that contains all the
  information about failed and successful contract assertions in the wrapped
  code blocks. This is useful for post-hoc investigation. Multiple type-checking
  failures can introduce a lot of noise in the program output. To alleviate this
  problem, \code{contractr.severity} option can be set to \code{'silence'}.
  This suppresses failure messages but \contractr still performs type-checking whose
  results can be explicitly obtained as a data frame from the
  \code{get\_contracts} API function. R suppresses printing of warnings on the
  terminal when a function issues too many warnings. This can hide type-checking
  failure messages. Setting \code{contractr.severity} flag to \code{'error'}
  turns the type-checking failure warnings to errors, which halts the program at
  the point of contract failure.

\end{itemize}

While \contractr's primary logic has been implemented in C++, we have not
observed a single segfault during our use of the package, either during the
development and sanity-checking phase, or during the evaluation phase.
We conclude the discussion of \contractr by drawing reader's attention to the
engineering effort it takes to package experimental ideas in the form of stable
and easy-to-use tools. R has an eclectic mix of features and design oddities
that make this even harder.
